---
title: "Hierarchical EHR: A 5 step approach for data management"
author: "Uchechukwu Ikeaba"
date: October 23, 2019
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Background & Motivation
1) Massive number of patient encounter, results in high amount of stored data.

2) The ability to transform clinical data into useful insights to improve patient's care is of increased necessity.

3) This presentation highlights 5 basic steps (analytical techniques) on a fictitious hierarchical dataset.

#### Data Description
1) Data consist of multiple doctor's delivering babies normally and also by cesarean section.

2) Demographic information are recorded for each mother. This includes Sex of baby, 
Mother's Race, Mother's house location (zip code) and Mother's age. 

3) Mother's undergo 2 procedures (code). Either a normal Delivery: (720, 721, 724, 726, 728, 729, 731, 733, 736, 738, 740, 741, 742, 744) or Delivery by C-section (740, 741, 742, 744)

#### Load working libraries
```{r}
library(readxl) # Used to load dataset
library(dplyr) # Used in Step 1 (Checks for duplicate rows) & Step 3 (Create derived variables from merged dataset)
library(sqldf) # Used in Step 2 (Merge both dataset) & 5 (Prepare hierarchical data for aggregation)
library(stringr) # Used in Step 3 (Create derived variables from merged dataset)
library(gmodels) # Used in Step 4 (QA derived variables from merged dataset (csection and delivery)

```

#### Import doctors data
```{r}
doctors <- read_excel("/Users/KABSTAT/Downloads/ProgrammerTest/doctors.xlsx")
str(doctors)

```


#### Import mothers data
```{r}
mothers <- read_excel("/Users/KABSTAT/Downloads/ProgrammerTest/mothers.xlsx")
str(mothers)

```

### Step 1
#### Check for duplicate rows (Mothers data set)
```{r}
mothers_dup <- duplicated(mothers)
table(mothers_dup)

```

#### How many times is a row duplicated (numdup > 1) Mothers data set
```{r}
dt <- aggregate(list(numdup=rep(1,nrow(mothers))), mothers, length)
# dt[order(dt$numdup), ]
tail(dt[order(dt$numdup), ])

```


#### Check for duplicate rows (Doctors data set)
```{r}
doctors_dup <- duplicated(doctors)
table(doctors_dup)

```

#### How many times is a row duplicated (numdup > 1) Doctors data set
```{r}
# How many times is a row duplicated
dt <- aggregate(list(numdup=rep(1,nrow(doctors))), doctors, length)
# dt[order(dt$numdup), ]
tail(dt[order(dt$numdup), ])

```

### Step 2
#### Merge both dataset (unique rows)
```{r}
doctors_unique <- unique(doctors)
mothers_unique <- unique(mothers)

library(sqldf)

## Left join 
leftjoin <- sqldf("SELECT *
              FROM doctors_unique
              LEFT JOIN mothers_unique USING(Mother_ID)")
head(leftjoin)

```

### Step 3 
#### Create derived variables from merged dataset
```{r}
leftjoin <- leftjoin %>% mutate(delivery_PR1 = ifelse(str_detect(PR1, c("720|721|724|726|728|729|731|733|736|738|740|741|742|744")), "yes", "no")) %>% mutate(delivery_PR2 = ifelse(str_detect(PR2,c("720|721|724|726|728|729|731|733|736|738|740|741|742|744")), "yes", "no")) %>% mutate(csection_PR1 = ifelse(str_detect(PR1, c("740|741|742|744")), "yes", "no")) %>% mutate(csection_PR2 = ifelse(str_detect(PR2, c("740|741|742|744")), "yes", "no"))

str(leftjoin)

```


#### Create derived variables from merged dataset (prepare variables for aggregate)
```{r}
leftjoin2 <- leftjoin %>% mutate(delivery = ifelse((delivery_PR1=="yes" | delivery_PR2=="yes"), 'yes', 'no')) %>% mutate(csection = ifelse((csection_PR1=="yes" | csection_PR2=="yes"), 'yes', 'no')) %>% mutate(black = ifelse((RACE=="B"), 'black', 'not black')) 

# str(leftjoin2)

```

### Step 4 
#### QA derived variables from merged dataset (csection and delivery)
```{r}
CrossTable(leftjoin2$csection, leftjoin2$delivery, prop.t=FALSE, prop.r=FALSE,
         prop.c=FALSE, prop.chisq = FALSE)

```

### Step 5 
#### Prepare hierarchical data for aggregation
```{r}
leftjoin3 <- leftjoin2 %>% mutate(delivery = ifelse((delivery_PR1=="yes" | delivery_PR2=="yes"), 1, 0)) %>% mutate(csection = ifelse((csection_PR1=="yes" | csection_PR2=="yes"), 1, 0)) %>% mutate(black = ifelse((RACE=="B"), 1, 0)) 

# str(leftjoin3)
```


#### Use SQL for data aggregation
```{r}
# write a full aggregation command, grouping by your specified columns
finaldata <- sqldf( "select Doctor_ID,
                    count(*) as aggregate,
		                sum(delivery) as sum_delivery, 
		                sum(csection) as sum_csection, 
		                sum(black) as sum_black 
             from leftjoin3 
             group by Doctor_ID" )

# print your result showing doctors with more deliveries
finaldata[order(finaldata$aggregate), ]

```

#### QA
```{r}
# write a full aggregation command, grouping by your specified columns
QA_finaldata <- sqldf( "select count(*) as physician_count,
                    avg(sum_delivery) as mean_delivery_norm, 
			              avg(sum_csection) as mean_c_section, 
                    avg(sum_black) as mean_black
             from finaldata" )

# print your result
QA_finaldata

```